The Impact of Seasonal Changes on Mortality Rates in Toronto

### 2024-09-27 paper/paper.qmd
Upload - Paper

---
title: "The Impact of Seasonal Changes on Mortality Rates in Toronto"
subtitle: "Death Registry Statistics in Toronto"
author: 
  - Doran Wang
thanks: "Code and data are available at: https://github.com/Wang20030509/Sta304-Term-Paper-1"
date: today
date-format: long
abstract: "This paper analyzes death toll data across four civic centers (Scarborough, North York, Toronto and Etobicoke)) from 2011 to 2023, with the goal of identifying seasonal patterns and regional differences in mortality, which is made available to the public from Open Data Toronto. Utilizing death registry data, we categorized the deaths by season—Winter, Spring, Summer, and Fall—and aggregated the death tolls across the years for each region. The analysis revealed notable variations in mortality trends across both seasons and civic centers. The results show that Winter consistently reported higher death tolls compared to other seasons across most regions. Using visual representations generated through ggplot2, the findings highlight significant regional disparities, which suggest possible influences of environmental or socio-economic factors contributing to seasonal mortality trends."
format: pdf
toc: true
toc-title: "Table of Contents"
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

This study examines the impact of seasonal changes on death tolls in
Toronto between 2011 and 2023, utilizing death registry data from Open
Data Toronto. By categorizing deaths by season—Winter, Spring, Summer,
and Fall—and analyzing regional variations across the city's four civic
centers (Scarborough, North York, Toronto, and Etobicoke), the paper
aims to uncover significant trends in mortality rates throughout the
year. Previous studies have shown that temperature plays a significant
role in seasonal mortality [@citeClimate] [@citeBivariate], with cold
winters and heat waves contributing to an increased death toll, but the
exact influence of socio-economic and environmental factors at the local
level remains under-explored [@citeSeasonality] [@citeSeasonal].

@@

The remainder of this paper is structured as follows. @sec-data
describes the dataset and cleaning methods used in the analysis. @sec-results 
presents the results and key findings, and @sec-discussion offers a discussion of the implications of these results
as well as limitations and potential next steps.

@@

The datasets used for this report are the latest delay statistics from the 2023 year, `Death Registry Data` [@deathRegistryData], which was published by the City Clerk's Office from `Open Data Toronto` [@rOpenDataToronto].

@@

* `ggplot2` [@citeggplot] 
* `dplyr` [@citedplyr] 
* `readr` [@citereadr] 
* `tibble` [@citetibble]

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-death_registry_data
#| tbl-cap: "Sample Data"
#| warning: false
kable(head(raw_data))

@@

* CIVIC_CENTRE: This column represents the civic center code, corresponding to one of Toronto’s four main civic centers: Scarborough (SC), North York (NY), Toronto (TO), and Etobicoke (ET). These civic centers are used to geographically categorize the death registrations.

* DEATH_LICENSES: This column records the number of deaths registered within a given month at each respective civic center. It represents the official count of death licenses issued, reflecting the mortality data across the city over different time periods.

* PLACE_OF_DEATH: This column provides information on where the death occurred. It indicates whether the death took place inside the City of Toronto or outside its city limits. This information can help to identify trends in deaths occurring within the jurisdiction of the city as opposed to those occurring elsewhere.

* TIME_PERIOD: This column captures the specific month and year during which the death licenses were registered. The time period is formatted as "YYYY-MM", allowing for temporal analysis of death registrations over time. The combination of month and year in this column enables the dataset to be used for tracking changes in death registration patterns on a monthly and yearly basis.

The death registry that @tbl-death_registry_data samples contains 947 entries in total. The dataset provides valuable insights into the geographic and temporal distribution of death registrations across Toronto, which is the main dataset used to analyze seasonal mortality patterns and potential regional disparities from 2011 to 2023.

@@

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-summerized_data
#| tbl-cap: "Summarized Data"
#| warning: false
#| tbl-subcap: ["Total Death toll by Season (2011-2023)", "Average Deaths by Month (2011-2023)","Death Toll by Place of Death Across 12 Months","Death Toll by Civic Centre Across 12 Months"]
#| layout-ncol: 2

kable(head(total_deaths), col.names = c("Season", "Civic Centre", "Death Toll"))

kable(head(avg_death), col.names = c("Month", "Mean of the Death Toll"))

kable(head(place_summary), col.names = c("Month", "Place of Death", "Death Toll"))

kable(head(civic_summary), col.names = c("Month", "Civic Centre", "Death Toll"))

@@

Our results are summarized in @fig-deaths
Something
Something
Something

```{r warning=FALSE}
#| label: fig-place_summary
#| fig-cap: Death Toll by Civic Centre Across 12 Months (2011-2023)
#| echo: false

# Create the line graph
ggplot(place_summary, aes(x = month, y = death_licenses, color = place_of_death, group = place_of_death)) +
  geom_line() +
  geom_point() +
  labs(title = "Death Licenses by Place of Death Across 12 Months",
       x = "Month",
       y = "Total Death Licenses",
       color = "Place of Death") +
  theme_minimal() +
  scale_x_discrete(limits = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
```

```{r warning=FALSE}
#| label: fig-civic_summary
#| fig-cap: Death Toll by Civic Centre Across 12 Months (2011-2023)
#| echo: false

# Create the line graph
ggplot(civic_summary, aes(x = month, y = death_licenses, color = civic_centre, group = civic_centre)) +
  geom_line() +
  geom_point() +
  labs(x = "Month",
       y = "Total Death Licenses",
       color = "civic_centre") +
  theme_minimal() +
  scale_x_discrete(limits = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
```

```{r warning=FALSE}
#| label: fig-deaths
#| fig-cap: Total Death Toll by Season and Region (2011-2023)
#| echo: false

@@

```{r warning=FALSE}
#| label: fig-Total_Deaths
#| fig-cap: Total Death toll by Season (2011-2023)
#| echo: false

# Plot the total deaths from 2011 to 2024
ggplot(total_deaths, aes(x = season, y = death_licenses, fill = season)) +
  geom_bar(stat = "identity", alpha = 1) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4", "red")) + 
  theme_minimal() +
  labs(x = "Session", y = "Total Death Toll", fill = "Season") +
  coord_flip()

@@

```{r warning=FALSE}
#| label: fig-avg_deaths_by_month
#| fig-cap: Average Death Toll by Month (2011-2023)
#| echo: false

# Plot the average deaths by month from 2011 to 2023
ggplot(avg_death, aes(x = factor(month), y = death_licenses, fill = factor(month))) +
  geom_bar(stat = "identity", alpha = 1) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme_minimal() +
  theme_light() +
  labs(x = "Month", y = "Average Death Toll", fill = "Month") 
```

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The
discussion is a chance to show off what you know and what you learnt
from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

@@

>>>>

---
title: "The Impact of Seasonal Changes on Mortality Rates in Toronto"
subtitle: "Death Registry Statistics in Toronto"
author: 
  - Doran Wang
thanks: "Code and data are available at: [https://github.com/Wang20030509/Sta304-Term-Paper-1](https://github.com/Wang20030509/Sta304-Term-Paper-1)"
date: today
date-format: long
abstract: "This paper analyzes death toll data across four civic centers (Scarborough, North York, Toronto and Etobicoke) from 2011 to 2024, with the goal of identifying seasonal patterns and regional differences in mortality, which is made available to the public from Open Data Toronto. Utilizing death registry data, we categorized the deaths by season—Winter, Spring, Summer, and Fall—and aggregated the death tolls across the years for each region. The analysis revealed notable variations in mortality trends across both seasons and civic centers. The results show that Winter consistently reported higher death tolls compared to other seasons across most regions. Using visual representations generated through ggplot2, the findings highlight significant regional disparities, which suggest possible influences of environmental or socio-economic factors contributing to seasonal mortality trends."
format: pdf
toc: true
toc-title: "Table of Contents"
number-sections: true
bibliography: references.bib
---

@@

This study examines the impact of seasonal changes on death tolls in
Toronto between 2011 and 2024, utilizing death registry data from Open
Data Toronto. By categorizing deaths by season—Winter, Spring, Summer,
and Fall—and analyzing regional variations across the city's four civic
centers (Scarborough, North York, Toronto, and Etobicoke), the paper
aims to uncover significant trends in mortality rates throughout the
year. Previous studies have shown that temperature plays a significant
role in seasonal mortality [@citeClimate] [@citeBivariate], with cold
winters and heat waves contributing to an increased death toll, but the
exact influence of socio-economic and environmental factors at the local
level remains under-explored [@citeSeasonality] [@citeSeasonal].

@@

The remainder of this paper is structured as follows. @sec-data
describes the dataset and cleaning methods used in the analysis.
@sec-results presents the results and key findings, and @sec-discussion
offers a discussion of the implications of these results.Then, @sec-conclusion will be the conclusion

@@

The datasets used for this report are the latest delay statistics from
the 2024 year, `Death Registry Data` [@deathRegistryData], which was
published by the City Clerk's Office from `Open Data Toronto`
[@rOpenDataToronto].

@@

-   `ggplot2` [@citeggplot]
-   `dplyr` [@citedplyr]
-   `readr` [@citereadr]
-   `tibble` [@citetibble]

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-death_registry_data
#| tbl-cap: "Sample Data"
#| warning: false
kable(head(raw_data))
```

@@

-   CIVIC_CENTRE: This column represents the civic center code,
    corresponding to one of Toronto’s four main civic centers:
    Scarborough (SC), North York (NY), Toronto (TO), and Etobicoke (ET).
    These civic centers are used to geographically categorize the death
    registrations.

-   DEATH_LICENSES: This column records the number of deaths registered
    within a given month at each respective civic center. It represents
    the official count of death licenses issued, reflecting the
    mortality data across the city over different time periods.

-   PLACE_OF_DEATH: This column provides information on where the death
    occurred. It indicates whether the death took place inside the City
    of Toronto or outside its city limits. This information can help to
    identify trends in deaths occurring within the jurisdiction of the
    city as opposed to those occurring elsewhere.

-   TIME_PERIOD: This column captures the specific month and year during
    which the death licenses were registered. The time period is
    formatted as "YYYY-MM", allowing for temporal analysis of death
    registrations over time. The combination of month and year in this
    column enables the dataset to be used for tracking changes in death
    registration patterns on a monthly and yearly basis.

The death registry that @tbl-death_registry_data samples contains 947
entries in total. The dataset provides valuable insights into the
geographic and temporal distribution of death registrations across
Toronto, which is the main dataset used to analyze seasonal mortality
patterns and potential regional disparities from 2011 to 2024.

@@

To prepare the Death Registry Statistics dataset for analysis, several
steps were taken to clean and standardize the data. These steps are
essential to ensure consistency and accuracy, enabling meaningful
comparisons across civic centers and time periods. The key cleaning
procedures applied to the raw data include:

1.  Renaming Columns: All column names were converted to more formal
    ones (e.g., death_licenses was renamed to Death Toll) for easier
    understanding in the table to the reader.

2.  Splitting the TIME_PERIOD Column: The time_period column, originally
    formatted as "YYYY-MM", was split into two separate columns: year
    and month. This allowed for more granular temporal analysis and
    easier data aggregation by season.

3.  Categorizing MONTH into Seasons: Based on the month column, a new
    season column was created to categorize each record into one of the
    four seasons:

    -   Winter (December, January, February)
    -   Spring (March, April, May)
    -   Summer (June, July, August)
    -   Fall (September, October, November)

4.  Standardizing PLACE_OF_DEATH: The place_of_death column, which
    indicates whether the death occurred within or outside Toronto, was
    standardized. The text was converted to title case to ensure
    consistency (e.g., "Outside City Limits" and "Toronto").

5.  Handling Missing or Invalid Data: Any missing or invalid entries in
    the dataset were either filled or removed where necessary to
    maintain the integrity of the analysis. For example, rows with
    missing or invalid death license counts were excluded from the final
    dataset.

After applying these cleaning procedures, the dataset was saved in its
cleaned form as cleaned_data.csv, and it is now ready for use in
subsequent analysis. This cleaned version of the data contains reliable
and well-structured information on death registrations by season and
civic center from 2011 to 2024.

@@

After cleaning and processing the Death Registry Statistics dataset, the
data was aggregated and summarized to provide insights into seasonal and
geographic trends in mortality across Toronto from 2011 to 2024. The
summary tables are Total Death Toll by Season and Civic Centre, Average
Deaths by Month, Death Toll by Place of Death, Death Toll by Civic
Centre. As shown in Table 3 and for more detailed
information, please refer to the [Appendix](#appendix). These summary
tables are primarily intended to facilitate the plotting and analysis in
the Results section.

@@

Here are 3 part of result with figures

## Distribution of deaths in different civic centers by season

```{r warning=FALSE}
#| label: fig-deaths
#| fig-cap: Total Death Toll by Season and Region (2011-2024)
#| echo: false

@@

As shown in @fig-deaths The total death toll in each season reveals
notable seasonal differences. Winter consistently reported higher death
tolls, particularly in Etobicoke (ET) and North York (NY), compared to
the other civic centers. This pattern suggests a potential link between
colder weather and increased mortality, aligning with previous studies
on the impact of temperature on health outcomes.

## Distribution of mean of deaths overall months

```{r warning=FALSE}
#| label: fig-avg_deaths_by_month
#| fig-cap: Average Death Toll by Month (2011-2024)
#| echo: false

# Plot the average deaths by month from 2011 to 2024
ggplot(avg_death, aes(x = factor(month), y = death_licenses, fill = factor(month))) +
  geom_bar(stat = "identity", alpha = 1) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme_minimal() +
  theme_light() +
  labs(x = "Month", y = "Average Death Toll", fill = "Month") 
```

As shown in @fig-avg_deaths_by_month January consistently reported the
highest average death toll, with a mean of 336 (from
Table 3, see more detail in [Appendix](#appendix))deaths.
This could be attributed to the colder winter months in Toronto, which
may exacerbate health conditions such as cardiovascular and respiratory
issues. In contrast, the summer months, such as July and August,
reported lower average death tolls.

## Total death toll overall seasons
```{r warning=FALSE}
#| label: fig-Total_Deaths
#| fig-cap: Total Death toll by Season (2011-2024)
#| echo: false

# Plot the total deaths from 2011 to 2024
ggplot(total_deaths, aes(x = season, y = death_licenses, fill = season)) +
  geom_bar(stat = "identity", alpha = 1) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4", "red")) + 
  theme_minimal() +
  labs(x = "Season", y = "Total Death Toll", fill = "Season") +
  coord_flip()

@@

We can see intuitively from @fig-Total_Deaths that the death toll in
winter and spring, two seasons with relatively low temperatures, is
higher than that in summer and fall.Also, there are more deaths in fall
than in summer, since the temperature gradually drops as fall begins.

## Death toll trends vary by season
```{r warning=FALSE}
#| label: fig-place_summary
#| fig-cap: Death Toll by Place of Death Across 12 Months (2011-2024)
#| echo: false

# Create the line graph
ggplot(place_summary, aes(x = month, y = death_licenses, color = place_of_death, group = place_of_death)) +
  geom_line() +
  geom_point() +
  labs(x = "Month",
       y = "Total Death Toll",
       color = "Place of Death") +
  theme_minimal() +
  scale_x_discrete(limits = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

```{r warning=FALSE}
#| label: fig-civic_summary
#| fig-cap: Death Toll by Civic Centre Across 12 Months (2011-2024)
#| echo: false

# Create the line graph
ggplot(civic_summary, aes(x = month, y = death_licenses, color = civic_centre, group = civic_centre)) +
  geom_line() +
  geom_point() +
  labs(x = "Month",
       y = "Total Death Toll",
       color = "Civic Centre") +
  theme_minimal() +
  scale_x_discrete(limits = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
```

As @fig-place_summary and @fig-civic_summary shows, regardless of
whether the death was found outside the city or in the Toronto area, the
death toll curve is lowest in the summer (June-August), and starts to
rise in the winter until it reaches its highest point in January. Even
when we distinguish by civic centre, the curve is similar and within
some errors.

@@

The analysis of mortality trends in Toronto from 2011 to 2024 reveals significant seasonal and regional variations. Winter consistently shows the highest death tolls across all civic centers, particularly in Etobicoke and North York. This finding aligns with established research linking colder temperatures to increased mortality, especially among vulnerable populations such as the elderly and those with pre-existing health conditions. Cardiovascular and respiratory diseases, exacerbated by cold weather, likely contribute to these elevated death tolls in the winter months.

Spring also exhibits relatively high mortality rates, possibly due to the lingering effects of cold temperatures and the transition between seasons. In contrast, summer consistently reports the lowest death tolls, likely due to milder temperatures and lower incidence of seasonal illnesses. Interestingly, fall shows an increase in death tolls as temperatures begin to drop, signaling the onset of winter-related health complications.

Regional disparities in mortality were evident, with Etobicoke consistently reporting the highest death tolls across all seasons. This may be attributed to socio-economic and environmental factors specific to the region, such as access to healthcare, population density, and the presence of vulnerable populations. North York also showed similar trends, while Toronto and Scarborough generally reported lower death tolls.

# Conclusion

This study provides valuable insights into the seasonal and regional patterns of mortality in Toronto over a 13-year period. The findings reinforce the significant impact of cold weather on mortality, particularly in winter and spring, and highlight the need for targeted public health interventions during these high-risk periods. The regional disparities observed suggest that localized factors play a crucial role in influencing mortality rates, indicating the need for region-specific strategies to reduce mortality.

@@

# Additional table details

As shown in @tbl-a, the dataset was grouped by season and civic center to examine the variations in mortality over time. As shown in @tbl-b, an analysis of the monthly average death toll reveals interesting patterns throughout the year. As shown in @tbl-c, the dataset was further analyzed to examine the location of death registrations. The summarized data by civic center (@tbl-d) shows that Etobicoke (ET) consistently recorded the highest death tolls in all months.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: tbl-a
#| tbl-cap: "Total Death toll by Season (2011-2024)"

kable(head(total_deaths), col.names = c("Season", "Civic Centre", "Death Toll"))
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: tbl-b
#| tbl-cap: "Average Deaths by Month (2011-2024)"

kable(head(avg_death), col.names = c("Month", "Mean of the Death Toll"))
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: tbl-c
#| tbl-cap: "Death Toll by Place of Death Across 12 Months"

kable(head(place_summary), col.names = c("Month", "Place of Death", "Death Toll"))
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: tbl-d
#| tbl-cap: "Death Toll by Civic Centre Across 12 Months"

kable(head(civic_summary), col.names = c("Month", "Civic Centre", "Death Toll"))
```
